Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis

### 2024-04-03 paper/paper.qmd
update intro
playing around with different figures for data section

Tumor size is often the main determinant of stage and treatment. As tumor categories increase, the tumor expands, invading nearby structures (Zhang, 2015). A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival. The authors found that patients presenting with larger tumors predicted a worse prognosis and thus are associated with a decrease in survival. There is a similar relationship between extensive lymph node involvement and patient survival (Zhang, 2015). Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones (Markman, 2023).  A study looked at five year survival rates based on the severity of spread. 62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis. More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023) . 
Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates (Zhang, 2015).

@@

Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases (Kai, 2021). In the early stage, cancer has not spread to lymph nodes. As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement  (Park, 2017). Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options (Kai, 2021). A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis (Park, 2017).

Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States. These patients were treatment naive and underwent surgical resection of the tumor. Although 5757 patients died, the rest showed great results, with a median survival of 22 months (Kai, 2021). The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios. A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival. The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival (Kai, 2021).

After tumor size, LN involvement and presence of distant metastasis are categorized, the pathogenic stage of the cancer is then determined (Eldridge, 2022). The most valuable prognostic factor in non-small cell lung cancer is the pathogenic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread. Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor (Eldridge, 2022). Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) (Park, 2017). A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages (Denton, 2016). The earlier the cancer is found, that is the lower the pathogenic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).

However, there is minimal literature looking at post-diagnosis survival rates based on pathogenic stage and method of treatment. The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods. Over the past decade, there has been great improvement of lymph node assessment in cancer patients (Kai, 2021). Experts hope survival rates continue to improve with  new therapies and treatment approaches  (Markman, 2023).

@@

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

>>>>

Tumor size is often the main determinant of stage and treatment. As tumor categories increase, the tumor expands, invading nearby structures (Zhang, 2015). A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival. The authors found that patients presenting with larger tumors predicted a worse prognosis and thus are associated with a decrease in survival. There is a similar relationship between extensive lymph node involvement and patient survival (Zhang, 2015). Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones (Markman, 2023).  A study looked at five year survival rates based on the severity of spread. 62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis. More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023). Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates (Zhang, 2015).

@@

After tumor size, LN involvement and presence of distant metastasis are categorized, the pathogenic stage of the cancer is then determined (Eldridge, 2022). The most valuable prognostic factor in non-small cell lung cancer is the pathogenic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread. Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor (Eldridge, 2022). Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) (Park, 2017). A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages (Denton, 2016). The earlier the cancer is found, that is the lower the pathogenic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022). However, there is minimal literature looking at post-diagnosis survival rates based on pathogenic stage and method of treatment. The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods. Over the past decade, there has been great improvement of lymph node assessment in cancer patients (Kai, 2021). Experts hope survival rates continue to improve with  new therapies and treatment approaches  (Markman, 2023).

@@

```{r}
#| label: fig-bills
#| fig-cap: bb
#| echo: false

# install.packages
# install.packages("ggplot2")
# install.packages("dplyr")

# load packaged
library(ggplot2)
library(dplyr)

# Histogram of lung cancer patients survival time in days, post-diagnosis
ggplot(analysis_data, aes(x = days_to_death)) +
  geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
  labs(title = "Distribution of Survial Time in Lung Cancer Patients Post-Diagnosis", x = "Survival Time (days)", y = "Frequency") + theme_minimal()

# create bar chart for pathogenic stage
pathogenic_stage_counts <- analysis_data %>%
  group_by(pathogenic_stage) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(pathogenic_stage))

ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  labs(title = "Percentage of Patients by Pathogenic Stage",
       x = "Pathogenic Stage",
       y = "Percentage") +
  theme_minimal()

# Create bar chart for treatment type
treatment_type_counts <- analysis_data %>%
  group_by(treatment_type) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(treatment_type))

ggplot(treatment_type_counts, aes(x = treatment_type, y = count)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  labs(title = "Distribution of Patients by Treatment Type",
       x = "Treatment Type",
       y = "Count") +
  theme_minimal()

ggplot(treatment_type_counts, aes(x = treatment_type, y = percentage)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  labs(title = "Percentage of Patients by Treatment Type",
       x = "Treatment Type",
       y = "Percentage") +
  theme_minimal()

# trying to represent percentage of pts by treatment type in a different way

# Calculate counts and percentages of patients by treatment type
treatment_type_counts <- analysis_data %>%
  group_by(treatment_type) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Pie chart for percentage of patients by treatment type
ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(title = "Percentage of Patients by Treatment Type",
       fill = "Treatment Type",
       x = NULL,
       y = NULL) +
  theme_void()

# Stacked bar chart for percentage of patients by treatment type
ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Patients by Treatment Type",
       x = NULL,
       y = "Percentage") +
  theme_minimal() +
  theme(legend.position = "bottom")

# dont like those either

# Donut chart for percentage of patients by treatment type
ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(title = "Percentage of Patients by Treatment Type",
       fill = "Treatment Type",
       x = NULL,
       y = NULL) +
  theme_void() +
  theme(legend.position = "right") +  # Adjust legend position if necessary
  annotate("text", x = 0, y = 0, label = "Total", size = 10, color = "black")  # Add label for total in the center

# hmm maybe donut was cuter

# Waffle chart for percentage of patients by treatment type
# install.packages("waffle")
library(waffle)

# Prepare data for waffle chart
waffle_data <- treatment_type_counts %>%
  mutate(squares = percentage / 2)  # Adjust scale if needed for better visualization

# Create waffle chart
waffle(waffle_data$squares, rows = 10, size = 0.6, colors = c("lightblue", "lightgreen"),
       title = "Percentage of Patients by Treatment Type", 
       xlab = "1 square = 2%", 
       pad = 1)  # Adjust padding between squares if necessary

# idea for the morning - sub in presence of distant metastasis, lymph node involvement and tumor size for pathogenic stage (frist graph)
